Supplementary figures for Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis

### 2020-12-03 manuscript/manuscript.Rmd
Tweaks from KF, TS suggestions

---
title: "Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
output: 
  #rticles::arxiv_article
  #pdf_document
  # bookdown::pdf_document2:
  #    toc: false
  bookdown::word_document2: 
    toc: false
    number_sections: false
    reference_docx: "output_template.docx"
abstract: |  
  **Abstract:** We collected a multi-centric retrospective dataset of patients (N = 213) who were admitted to ten hospitals in Czech Republic and tested positive for SARS-CoV-2. The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution. Patients in the dataset were treated with hydroxychloroquine (N = 108), azithromycin (N = 72), favipiravir (N = 9), convalescent plasma (N = 7), dexamethasone (N = 4) and remdesivir (N = 3), often in combination. Most patients were admitted during the first wave of the epidemic. To explore association between treatments and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom.  Weak evidence to constrain the potential efficacy of azithromycin and favipiravir can be extracted from the data. Additionally, we performed external validation of several proposed prognostic models for Covid-19 severity showing that they mostly perform unsatisfactorily on our dataset.
bibliography: Covid19.json
csl: journal-of-medical-virology.csl
date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
author: "`r { source(here::here('manuscript', 'authors.R')); get_authors_string() }`"
---

@@

The Covid-19 pandemic caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of November 2020, led to over 46 million cases and over 1.1 million deaths. Proposed treatments include antivirals approved for other indications (chloroquine, hydroxychloroquine, lopinavir/ritonavir, remdesivir, favipiravir, umifenovir), azithromycin, corticosteroids, immunoglobulins, tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR]. As of this writing, the only treatment that exhibited positive effects in randomized trials is the corticosteroid dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58]. Remdesivir, previously designed and approved for Ebola, SARS and MERS, is currently the only antiviral against SARS-CoV-2 approved by the FDA. However, during spring and summer it was available only in limited amounts via compassionate use programme. Randomized clinical trials on remdesivir have provided  conflicting evidence [@http://zotero.org/users/5567156/items/MFS3IPPU; @http://zotero.org/users/5567156/items/KB5UDZH3]. Despite initial promise, hydroxychloroquine has failed to show benefit for hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM; @http://zotero.org/users/5567156/items/KB5UDZH3], as an early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], as well as pre- and post-exposure prophylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR]. Further, no improvement could be discerned when adding azithromycin to hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ]. Similarly, Lopinavir/Ritonavir did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU]. No benefit was observed for convalescent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; @http://zotero.org/users/5567156/items/BHGT88US].

@@

Favipiravir is one of the less explored experimental treatments, previously used for influenza including avian types, with only one larger RCT in Covid- 19 known to us [@http://zotero.org/users/5567156/items/ISLSL7UC], the study however has many potentially severe problems [@http://zotero.org/users/5567156/items/7HDXLWEA]. Other studies are scarce - even when considering very small studies and case series [@http://zotero.org/users/5567156/items/F3VNTQ92; @http://zotero.org/users/5567156/items/FSUQMUP8; @http://zotero.org/users/5567156/items/894H9N9B]. None of the results allow strong conclusions about benefits or harms of favipiravir, although a  metaanalysis of all available small studies showed an effect of favipiravir on clinical improvement in mild and moderate disease [@http://zotero.org/users/5567156/items/7595KBWR].

 For additional summary of treatments under investigation see Siemieniuk et al. [@http://zotero.org/users/5567156/items/YP62P6XJ]

@@

In total, we were able to gather data for `r n_patients` patients, see Table \@ref(tab:summarytable) for the overall characteristics of the patient sample and several subgroups we used in the analysis, including treatments taken. The dataset includes 19 patients already reported in a study of inflammatory signatures of Covid-19 [@http://zotero.org/users/5567156/items/IA4FQCF3].

@@

fig_progression <- data_for_fig_progression %>% 
  filter(!is_outcome) %>%
  ggplot(aes(x = day, y = patient_id, fill = breathing_s)) + 
  geom_tile(height = 0.7) + 
  geom_point(data = data_for_fig_progression %>% filter(is_outcome) %>% mutate(outcome = factor(breathing_s, ordered = FALSE)),
             inherit.aes = FALSE, aes(x = day, y = patient_id, shape = outcome), size = 0.9) +
  scale_fill_viridis_d("", direction = -1) +
  scale_x_continuous("Days since hospitalization") +
  scale_shape_manual("", values = c(1, 15), guide = guide_legend(override.aes = list(size = 4))) +
  facet_wrap(~first_half, scales = "free_y") +
  theme(axis.text.y = element_blank(), strip.background = element_blank(), strip.text = element_blank(),
        legend.direction = "horizontal", legend.position = "bottom") +
  theme(axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank())

@@

As noted above, the nature of the convenience sample did not enforce random assignment of treatments to patients. In fact, patients with worse baseline characteristics, which lead to worse outcomes, were less likely to receive hydroxychloroquine (see Supplementary Figure 2). This clearly creates a bias towards a positive effect of hydroxychloroquine on the outcome (and potentially for other treatments as well - most were used in combination with hydroxychloroquine), which, however, could be false.

@@

first_fig_multiverse_drugs <- 4
last_fig_multiverse_drugs <- 6
supplementary_figure_markers <- last_fig_multiverse_drugs + 1 
supplementary_figures_multiverse <-  paste0(first_fig_multiverse_drugs, "-", last_fig_multiverse_drugs)

@@

Here, we focus on hydroxychloroquine and azithromycin as those are the only treatments with larger sample size. We also investigate favipiravir as it is less well reported in the literature. Hydroxychloroquine was dosed almost exclusively in a 5-day regime starting with a loading dose of 800mg on the first day and followed by 400mg. Majority of patients complemented hydroxychloroquine with azithromycin while azithromycin was rarely used alone (see Table \@ref(tab:summarytable)). Azithromycin was most frequently dosed 250 or 500mg/day, but doses ranging from 100mg/day to 1500 mg/day were observed.  Favipiravir was used only at one site with a loading dose of 3600mg on the first day, followed by at most 9 days with a 1600mg dose. All but one of the patients receiving favipiravir also received hydroxychloroquine. Treatment was initiated mostly within two days of admission (see Supplementary Figure 3).

>>>>

---
title: "Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
output: 
  #rticles::arxiv_article
  #pdf_document
  # bookdown::pdf_document2:
  #    toc: false
  bookdown::word_document2: 
    toc: false
    number_sections: false
    reference_docx: "output_template.docx"
abstract: |  
  **Abstract:** We collected a multi-centric retrospective dataset of patients (N = 213) who were admitted to ten hospitals in Czech Republic and tested positive for SARS-CoV-2. The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution. Patients in the dataset were treated with hydroxychloroquine (N = 108), azithromycin (N = 72), favipiravir (N = 9), convalescent plasma (N = 7), dexamethasone (N = 4) and remdesivir (N = 3), often in combination. Most patients were admitted during the first wave of the epidemic. To explore association between treatments and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom.  Weak evidence to constrain the potential efficacy of azithromycin and favipiravir can be extracted from the data. Additionally, we performed external validation of several proposed prognostic models for Covid-19 severity showing that they mostly perform unsatisfactorily on our dataset.
bibliography: Covid19.json
csl: journal-of-medical-virology.csl
#date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
author: "`r { source(here::here('manuscript', 'authors.R')); get_authors_string() }`"
---

@@

The Covid-19 pandemic caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of November 2020, led to over 46 million cases and over 1.1 million deaths. Proposed treatments include antivirals approved for other indications (chloroquine, hydroxychloroquine, lopinavir/ritonavir, remdesivir, favipiravir, umifenovir), azithromycin, corticosteroids, immunoglobulins, tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR]. As of this writing, the only treatment that exhibited positive effects in randomized trials is the corticosteroid dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58]. Remdesivir, previously designed and approved for Ebola, SARS and MERS, is currently the only antiviral against SARS-CoV-2 approved by the FDA. However, during spring and summer it was available only in limited amounts via compassionate use programme. Randomized clinical trials on remdesivir have provided  conflicting evidence [@http://zotero.org/users/5567156/items/MFS3IPPU; @http://zotero.org/users/5567156/items/KB5UDZH3]. Despite initial promise, hydroxychloroquine has failed to show benefit for hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM; @http://zotero.org/users/5567156/items/KB5UDZH3], as an early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], as well as pre- and post-exposure prophylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR]. Further, no improvement could be discerned when adding azithromycin to hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ]. Similarly, lopinavir/ritonavir did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU]. No benefit was observed for convalescent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; @http://zotero.org/users/5567156/items/BHGT88US].

@@

Favipiravir is one of the less explored experimental treatments, previously used for influenza including avian types, with only one larger RCT in Covid- 19 known to us [@http://zotero.org/users/5567156/items/ISLSL7UC], the study however has many potentially severe problems [@http://zotero.org/users/5567156/items/7HDXLWEA]. Other studies are scarce - even when considering very small studies and case series [@http://zotero.org/users/5567156/items/F3VNTQ92; @http://zotero.org/users/5567156/items/FSUQMUP8; @http://zotero.org/users/5567156/items/894H9N9B]. None of the results allow strong conclusions about benefits or harms of favipiravir, although a  metaanalysis of all available small studies showed an effect of favipiravir on clinical improvement in mild and moderate disease [@http://zotero.org/users/5567156/items/7595KBWR].
For additional summary of treatments under investigation see Siemieniuk et al. [@http://zotero.org/users/5567156/items/YP62P6XJ]

@@

In total, we were able to gather data for `r n_patients` patients, see Table \@ref(tab:summarytable) for the overall characteristics of the patient sample and several subgroups we used in the analysis, including treatments taken. 
Counts of all treatment combinations are shown in Supplementary Figure 2 and Supplementary Figure 3 shows outcomes by study site, demonstrating quite large hospital-specific differences.
The dataset includes 19 patients already reported in a study of inflammatory signatures of Covid-19 [@http://zotero.org/users/5567156/items/IA4FQCF3].

@@

fig_progression <- data_for_fig_progression %>% 
  filter(!is_outcome) %>%
  ggplot(aes(x = day, y = patient_id, fill = breathing_s)) + 
  geom_tile(height = 0.7) + 
  geom_point(data = data_for_fig_progression %>% filter(is_outcome) %>% mutate(outcome = factor(breathing_s, ordered = FALSE)),
             inherit.aes = FALSE, aes(x = day, y = patient_id, shape = outcome), size = 0.9) +
  scale_fill_viridis_d("", direction = -1) +
  scale_x_continuous("Days since hospitalization") +
  # scale_color_manual("", values = c("black", "darkred"))+
  scale_shape_manual("", values = c(1, 15), guide = guide_legend(override.aes = list(size = 4))) +
  facet_wrap(~first_half, scales = "free_y") +
  theme(axis.text.y = element_blank(), strip.background = element_blank(), strip.text = element_blank(),
        legend.direction = "horizontal", legend.position = "bottom") +
  theme(axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank())

@@

As noted above, the nature of the convenience sample did not enforce random assignment of treatments to patients. In fact, patients with worse baseline characteristics, which lead to worse outcomes, were less likely to receive hydroxychloroquine (see Supplementary Figure 4). This clearly creates a bias towards a positive effect of hydroxychloroquine on the outcome (and potentially for other treatments as well - most were used in combination with hydroxychloroquine), which, however, could be false.

@@

first_fig_multiverse_drugs <- 6
last_fig_multiverse_drugs <- 8
supplementary_figure_markers <- last_fig_multiverse_drugs + 1 
supplementary_figures_multiverse <-  paste0(first_fig_multiverse_drugs, "-", last_fig_multiverse_drugs)

@@

Here, we focus on hydroxychloroquine and azithromycin as those are the only treatments with larger sample size. We also investigate favipiravir as it is less well reported in the literature. Hydroxychloroquine was dosed almost exclusively in a 5-day regime starting with a loading dose of 800mg on the first day and followed by 400mg. Majority of patients complemented hydroxychloroquine with azithromycin while azithromycin was rarely used alone (see Table \@ref(tab:summarytable)). Azithromycin was most frequently dosed 250 or 500mg/day, but doses ranging from 100mg/day to 1500 mg/day were observed.  Favipiravir was used only at one site with a loading dose of 3600mg on the first day, followed by at most 9 days with a 1600mg dose. All but one of the patients receiving favipiravir also received hydroxychloroquine. Treatment was initiated mostly within two days of admission (see Supplementary Figure 5).

### 2020-12-03 manuscript/supplementary_figures.Rmd
Tweaks from KF, TS suggestions

devtools::load_all()
library(tidyverse)
library(patchwork)
library(tsiMisc) # devtools::install_github("tsieger/tsiMisc")

@@

@@

marker_map <- c("az" = "Azithromycin", "hcq" = "Hydroxychloroquine", "favipiravir" = "Favipiravir", "convalescent_plasma" = "Conv. plasma")

data$marker_data %>% 
  filter(marker %in% c("az", "hcq", "favipiravir"), !is.na(value), value > 0) %>%
  mutate(marker = marker_map[marker]) %>%
  group_by(marker, patient_id) %>% 
  summarise(first_day = min(day), .groups = "drop") %>% 
  ggplot(aes(x = first_day)) + geom_histogram(binwidth = 1) + facet_wrap(~marker, scales = "free", nrow = 1) + scale_x_continuous("First day of treatment relative to admission")
```

>>>>

devtools::load_all()
library(tidyverse)
library(patchwork)
library(tsiMisc) # devtools::install_github("tsieger/tsiMisc")
library(UpSetR)

@@

```{r upsettreatment, fig.cap = "Upset plot of treatment combinations - each vertical bar displays the number of patients that received the combination indicated by filled dots in the matrix. Horizontal bars show the total number of patients receiveing the given treatment.", fig.width=8, fig.height=6 }
upset_data <- data$patient_data %>% select(starts_with("ever")) %>% mutate(across(starts_with("ever"), as.integer)) %>% as.data.frame()

names(upset_data) <- gsub("ever_", "", names(upset_data)) %>% gsub("_", " ", .) %>% gsub("hcq", "hydroxychloroquine", .) %>% gsub("az", "azithromycin", .)

names(upset_data) <- paste0(toupper(substring(names(upset_data),1,1)), substring(names(upset_data),2))

UpSetR::upset(upset_data, sets = names(upset_data), text.scale = 2)
```

```{r outcome_by_hospital_prepare, echo=FALSE}
tmp<-data$patient_data[,c('patient_id','hospital_id','outcome')]
tmp$outcome[tmp$outcome=='Transferred']<-'Hospitalized'
tmp$outcome<-dropLevels(tmp$outcome)
tmp$hospital_id<-factor(tmp$hospital_id)
levels(tmp$hospital_id)<-paste('H',1:length(levels(tmp$hospital_id)),sep='')
```

```{r outcomebyhospital, echo=FALSE}
# knitr::kable(with(tmp,table(outcome,hospital_id)),caption = paste0( "Overview of outcomes in different hospitals. Note that at H6, 9 out of 19 patients died, while at H9, no patient died out of 9 patients treated. There were large differences between hospitals in the ratio of fatal outcomes (P=", round(fisher.test(with(tmp,table(outcome,hospital_id)),simulate.p.value=T)$p.value,5), ")"), booktabs = TRUE)
```

```{r hospitaloutcomes,fig.cap="Number of patients and outcomes at the individual sites. The numbers above bars are the exact counts. Hospitalized = still hospitalized at the end of data collection at the site. Sites are anonymized to preserve patient privacy."}
data$patient_data %>% mutate(
  hospital_label = factor(paste0("H", as.integer(factor(hospital_id))), levels = paste0("H", 1:10)),
  outcome = fct_collapse(outcome, Hospitalized = c("Hospitalized", "Transferred")) %>% fct_relevel("Death", "Hospitalized", "Discharged")
) %>%
  ggplot(aes(x = hospital_label,  fill = outcome)) + geom_bar(position = position_dodge()) + 
  geom_text(stat='count', aes(label=..count..), vjust=-1, position = position_dodge(width = 1)) +
  #scale_fill_viridis_d("Outcome") +
  scale_fill_manual("Outcome", values = c(Death = as.character(bfc()["Black"]), Hospitalized = as.character(bfc()["Bluish green"]), Discharged = as.character(bfc()["Yellow"]))) +
  scale_x_discrete("Study site") +
  scale_y_continuous("Count") +
  expand_limits(y = 36)
```

@@

marker_map <- c("hcq" = "Hydroxychloroquine", "az" = "Azithromycin", "favipiravir" = "Favipiravir", "convalescent_plasma" = "Conv. plasma")

data$marker_data %>% 
  filter(marker %in% c("az", "hcq", "favipiravir"), !is.na(value), value > 0) %>%
  mutate(marker = factor(marker_map[marker], levels = marker_map)) %>%
  group_by(marker, patient_id) %>% 
  summarise(first_day = min(day), .groups = "drop") %>% 
  ggplot(aes(x = first_day)) + geom_histogram(binwidth = 1) + facet_wrap(~marker, scales = "free_y", nrow = 1) + scale_x_continuous("First day of treatment relative to admission")
```
